---
title: "Wine Ratings by Variety and Origin"
author: "John Donnellan"
date: "2026-02-18"
categories: [analysis]
---

**Introduction:**

This analysis uses a wine review dataset obtained from Kaggle, which contains detailed reviews of wines from various countries, and grape varieties around the world. After cleaning the data, the dataset includes over 12,000 observations, each with information on the wine’s origin, grape variety, price, and rating.

How does the country of origin, variety of grape, and price affect the overall rating of the wine?

Link to data set: https://www.kaggle.com/datasets/zynicide/wine-reviews

**The Variables:**

-   country: Country of origin

-   points: Number of points WineEnthusiast rated the wine on a scale of 1-100

-   price: The cost for a bottle of the wine

-   variety: Grape type

**Loading and Cleaning the Data:**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(readr)
wine_ratings <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-05-28/winemag-data-130k-v2.csv")

wine <- wine_ratings |>
  filter(!is.na(price)) |>
  filter(!is.na(country)) |>
  filter(!is.na(variety)) |>
  filter(!is.na(province)) |>
  select(country, points, price, variety, province)|>
  arrange(desc(price)) |>
  slice(-(1:100))
```

The dataset was first cleaned by removing observations with missing values for price, country, variety, and province to ensure that all variables used in the analysis were complete. We then selected only the relevant variables and removed the 100 most expensive wines, as these extreme price outliers were skewing the mean price and distorting comparisons across countries and varieties.

**Graph #1**

```{r}
wine |>
  group_by(variety) |>
  filter(n() > 3000) |>
  summarize(
    avg_price = mean(price),
    avg_points = mean(points)
  ) |>
  ggplot(aes(x = avg_price, y = avg_points, color = variety)) +
  geom_point() +
  labs(
    title = "Average Price vs Average Rating by Grape Variety",
    subtitle = "Includes grape varieties with more than 3,000 reviews",
    x = "Average Price",
    y = "Average Rating",
    color = "Grape Variety"
  ) +
  theme_minimal() 
```

Interpretation:

This scatterplot shows a positive relationship between average price and average rating across grape varieties, though the relationship is modest. Higher priced varieties such as Pinot Noir and Syrah tend to have higher average ratings, clustered near the top of the graph. In contrast, lower priced wines like Rosé and Merlot sit toward the lower end of the rating scale, though the differences in average points are smaller overall.

**Graph #2**

```{r}
overall_avg <- mean(wine$points)

wine |>
  group_by(country) |>
  filter(n() > 400) |>
  mutate(avg_price = mean(price),
         avg_points = mean(points)) |>
  ungroup()|>
  mutate(country = fct_reorder(country, avg_points)) |>
  ggplot(aes(x = country, y = points, fill = avg_price)) +
  geom_boxplot()+
  geom_hline(yintercept = overall_avg, linetype = "dashed")+
  scale_fill_viridis_b() +
  labs(
    title = "Wine Ratings by Country",
    subtitle = "Countries with More Than 400 Wines",
    x = "Country",
    y = "Points",
    fill = "Average Price") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
  
```

Interpretation:

This boxplot compares the distribution of wine ratings across countries with more than 400 reviewed wines, with color indicating each country’s average price binned into \$5 increments. While there is some variation in median ratings, most countries cluster between roughly 86 and 90 points, suggesting that overall quality is relatively consistent among countries. Countries with higher average prices do not always have higher average ratings, indicating that price alone does not guarantee more points.

**Graph #3**

```{r}
overall_avg <- mean(wine$points)

wine |>
  group_by(variety) |>
  filter(n() > 1200) |>
  mutate(
    avg_points = mean(points),
    avg_price = mean(price)) |>
  ungroup() |>
  mutate(variety = fct_reorder(variety, avg_points)) |>
  ggplot(aes(x = points, y = variety, fill = avg_price)) +
  geom_boxplot() +
  geom_vline(xintercept = overall_avg, linetype = "dashed") +
  scale_fill_viridis_c() +
  labs(
    title = "Wine Ratings by Grape Variety",
    subtitle = "Dashed line shows overall average rating",
    x = "Points",
    y = "Grape Variety",
    fill = "Average Price"
  ) +
  theme_minimal()
```

Interpretation:

This boxplot shows the distribution of ratings for grape varieties with more than 1,200 reviews, ordered by their average rating and colored by average price. Varieties such as Nebbiolo and Champagne Blend appear above the overall average rating (dashed line), indicating stronger ratings for expensive varieties. In contrast, varieties like Rose, Merlot, and White Blend tend to fall below the overall mean, though their distributions still overlap heavily with higher-rated types.

**Conclusion and Wrap-Up**

Overall, price, country of origin, and grape variety all show some relationship with wine ratings, but none appear to strongly determine quality on their own. Higher-priced wines and certain varieties tend to receive slightly higher average ratings, though the differences are relatively small. One limitation of this analysis is the removal of the 100 most expensive wines, which may exclude meaningful high-end observations while reducing skew.

**Connection to Class Ideas**

These visualizations apply key ideas from class about effective data communication. Ordering categories by average rating improves clarity and makes comparisons easier, while boxplots display the variation rather than relying only on means. The reference lines provide context for interpreting above and below average ratings. Additionally, color encoding allows an extra variable (average price) to be shown without making the visualizing.
